---
title: "Power analysis: optimisation of sea lice monitoring programme"
subtitle: "Abundance"
author: "Bela Klimesova"
date: "2026-06-16"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn=-1)
```

# Preparation of sea lice dataset

R packages necessary of data analysis
```{r}
library(ggplot2)
library(glmmTMB)
library(progress)
suppressMessages(library(dplyr))
suppressMessages(library(future.apply))
library(progressr)
```

## Sea lice dataset

Sea lice monitoring dataset provided by the industry producer
```{r}
Industry <- read.table(
  "Industry.csv",
  header = TRUE,
  sep = ',',
  fill = TRUE,
  dec = ".",
  na.strings = "NA"
)
```

# Empirical data analysis

## Modelling of temporal increase of sea lice abundance levels

Adjustment of the variable pen: unique identification for each farm
```{r}
Industry$Pen_un <- paste(Industry$SiteName, Industry$Pen.Count, sep = "_")
```

GLMM model
abundance increase per production week, random effects to take into account the hierarchical structure of the data

```{r}
mod_A <- glmmTMB(
  Adult.female ~ Week_Production + (1 | SiteName) + (1 | Pen_un) + (1 | Year_Production),
  family = nbinom2,
  data = Industry
)
```

Exploration of the model

```{r}
# Results overview
summary(mod_A)

# AIC
AIC(mod_A)

# Collinearity
performance::model_performance(mod_A)
```

## Exploration of temporal increases in infection levels

### Weekly increase for all farms
Rates ratio increase of abundance converted to %
```{r}
slope <- fixef(mod_A)$cond["Week_Production"]
RR <- exp(slope)
per_increase <- (RR - 1) * 100
per_increase
```

### Weekly increase for each farm separately

GLMM of temporal increase for each farm
```{r}
mod_A1 <- glmmTMB(
  Adult.female ~ Week_Production + (Week_Production |
                                             SiteName) +
    (1 | Pen_un) + (1 | Year_Production),
  family = nbinom2,
  data = Industry
)
```

Calculation of percentage increase, rates ratio increase of abundance converted to %
```{r}
beta_fixed <- fixef(mod_A1)$cond["Week_Production"]
ran <- ranef(mod_A1)$cond$SiteName
site_dev <- ran[, "Week_Production"]

site_slopes <- beta_fixed + site_dev
RR_sites <- exp(site_slopes)
perc_inc_sites <- (RR_sites - 1) * 100

results <- data.frame(
  Site = rownames(ran),
  Slope = site_slopes,
  Rate_Ratio = RR_sites,
  Percent_Increase = perc_inc_sites
)

results
```

### Abundance variation between two consecutive inspections

Comparing abundance levels between two consecutive visits in raw data

Abundance at farm level
```{r}
Ind1 <- aggregate(
  Industry$Adult.female,
  list(
    Industry$SiteName,
    Industry$Year_Production,
    Industry$Week_Production,
    Industry$Count.Date
  ),
  mean
)
colnames(Ind1) <- c('Site', 'Year', 'Week', 'Date', 'Abun')

Ind1$Date <- as.Date(Ind1$Date, format = "%Y-%m-%d")
Ind1_s <- Ind1[order(Ind1$Date), ]
```

Calculation of percentage change in abundance levels
```{r}
Ind1_s <- Ind1_s %>%
  arrange(Site, Date) %>%
  group_by(Site) %>%
  mutate(Abun_lag = lag(Abun),
         Increase = ifelse(Abun_lag == 0 | is.na(Abun_lag), NA,
                           ((Abun - Abun_lag) / Abun_lag) * 100)) %>% 
      ungroup()

Ind1_s$Increase <- round(Ind1_s$Increase)
```

Exploration of abundance change

```{r}
Ind1_s1 <- as.data.frame(Ind1_s)

# Increase
summary(Ind1_s1[Ind1_s1$Increase>0,]$Increase)

# Decrease
summary(Ind1_s1[Ind1_s1$Increase<0,]$Increase)
```

Increase in abundance from 0
```{r}
Ind_0<-Ind1_s[Ind1_s$Abun_lag == 0, ]
Ind_0 <- as.data.frame(Ind_0)
Ind_0<-Ind_0[!(is.na(Ind_0$Year)),]

nrow(Ind_0[Ind_0$Abun>0.19,])/nrow(Ind_0)
```

## Estimated parameters from GLMM

Slope: rate of abundance increase
```{r}
slope <- fixef(mod_A)$cond["Week_Production"]
```

Mean0: intercept on the log scale
```{r}
intercept <- fixef(mod_A)$cond["(Intercept)"]
mean0 <- exp(intercept)
```

Theta
```{r}
theta <- 1 / (sigma(mod_A)^2)
```

Random effects sd
```{r}
vc <- VarCorr(mod_A)$cond

site_sd <- as.numeric(attr(vc$SiteName, "stddev"))
pen_sd <- as.numeric(attr(vc$Pen_un, "stddev"))
year_sd <- as.numeric(attr(vc$Year_Production, "stddev"))
```

Number of fish per pen
```{r}
fish_per_pen <- Industry %>%
  group_by(SiteName, Date, Pen.ID) %>%
  summarise(n_fish = n_distinct(FishID), .groups = "drop") %>%
  summarise(mean_fish = mean(n_fish)) %>%
  pull(mean_fish) %>%
  round()
```

Number of sites and years
```{r}
n_sites <- length(unique(Industry$SiteName))
n_years <- length(unique(Industry$Year_Production))
```

# Simulation-based power analysis on national level

## Power analysis function

Function for simulation-based power analysis
```{r}
simulate_test_s <- function(site_sd,
                            pen_sd,
                            year_sd,
                            n_sites,
                            n_pens,
                            n_years,
                            fish_per_pen,
                            n_visits,
                            mean0,
                            slope,
                            theta,
                            n_reps = 100) {
  intercept <- log(mean0) # Convert baseline abundance to log scale
  
  # Run simulation
  power_results <- replicate(n_reps, {    # run the simulation n_reps times
    
    # random effect
    site_re <- rnorm(n_sites, 0, site_sd)
    pen_re  <- rnorm(n_sites * n_pens, 0, pen_sd)
    year_re <- rnorm(n_years, 0, year_sd)
    
    # study design
    dat <- expand.grid(
      Site = factor(1:n_sites),
      Pen = factor(1:n_pens),
      Week = 1:n_visits,
      Year = factor(1:n_years)
    )
    
    # random effect
    dat$Pen_un <- interaction(dat$Site, dat$Pen)
    pen_levels <- interaction(dat$Site, dat$Pen)
    
    dat$site_re <- site_re[dat$Site]
    dat$pen_re <- pen_re[as.numeric(pen_levels)]
    dat$year_re <- year_re[dat$Year]
      
    # Expand to fish-level data
    dat_fish <- dat[rep(seq_len(nrow(dat)), each = fish_per_pen), ]
    dat_fish$FishID <- rep(1:fish_per_pen, times = nrow(dat))
    
    # Linear predictor
    eta <- intercept + slope * dat_fish$Week +
      dat_fish$site_re + dat_fish$pen_re + dat_fish$year_re
    mu <- exp(eta)
    
    # Generate counts per fish
    dat_fish$Adult.female <- rnbinom(nrow(dat_fish), mu = mu, size = theta)
    
    # Fit model
    formula <- "Adult.female ~ Week"
    
    # Add random effects only if more than 1 level exists
    if (n_sites > 1)
      formula <- paste0(formula, " + (1|Site)")
    if (n_pens > 1)
      formula <- paste0(formula, " + (1|Pen_un)")
    if (n_years > 1)
      formula <- paste0(formula, " + (1|Year)")
    
    m <- try(glmmTMB(as.formula(formula), family = nbinom2, data = dat_fish),
             silent = TRUE)
    
    if (inherits(m, "try-error"))
      return(NA)	# If model fails to converge, skip this replicate
    if (!m$sdr$pdHess)
      return(NA) # skip non-converged models
    
    # Detection rule
    est <- summary(m)$coefficients$cond["Week", ]
    est["Estimate"] > 0 && est["Pr(>|z|)"] < 0.05
  })
  
  # Remove NAs (failed fits)
  power_clean <- power_results[!is.na(power_results)]
  
  # Calculate power and confidence interval
  power_estimate <- mean(power_clean)
  ci <- binom.test(sum(power_clean), length(power_clean))$conf.int
  
  list(
    power = power_estimate,
    lower_CI = ci[1],
    upper_CI = ci[2],
    n_reps = length(power_clean)
  )
}
```

## Sampling scenarios

Rates ratios (effect size)
```{r}
rate_ratios <- c(1.04, 1.06, 1.1, 1.2, 1.3)
```

Number of fish per pen
```{r}
fish_vec <- c(5, 10, 15, 20, 25, 30)
```

Number of visits
```{r}
visit_vec <- c(2, 3, 4, 6, 8)
```

Number of pens
```{r}
pen_vec <- c(2, 5, 8, 10, 12, 15)
```

Replicates
```{r}
n_reps <- 1000
```

## Simulation run

In the following text the code to run simulation is provided, however, results are uploaded separately.

Create design grid for all scenarios
```{r}
design_grid <- expand.grid(
  RR = rate_ratios,
  fish_per_pen = fish_vec,
  n_visits = visit_vec,
  n_pens = pen_vec
)
```

Prepare columns for results
```{r}
design_grid$power <- NA_real_
design_grid$lower_CI <- NA_real_
design_grid$upper_CI <- NA_real_
design_grid$n_success <- NA_integer_
```

Set parallel backend
```{r}
# plan(multisession) 
```

Reproducibility
```{r}
# set.seed(123)
```

Progress bar
```{r}
# handlers(global = TRUE)
# handlers("txtprogressbar") 
```

Simulation
```{r}
# with_progress({
#
#  p <- progressr::progressor(steps = nrow(design_grid))

#  results_list <- future_lapply(seq_len(nrow(design_grid)), function(i) {
  
  # Convert RR to slope
#  slope_i <- log(design_grid$RR[i])
  
  # Run simulation
# sim_res <- simulate_test_s(
#    site_sd = site_sd,
#    pen_sd = pen_sd,
#    year_sd = year_sd,
#    n_sites = n_sites,
#    n_pens = design_grid$n_pens[i],
#    n_years = n_years,
#    fish_per_pen = design_grid$fish_per_pen[i],
#    n_visits = design_grid$n_visits[i],
#    mean0 = mean0,
#    slope = slope_i,
#    theta = theta,
#    n_reps = n_reps
#  )

#    p()  
  
  # Return results
#  list(
#    power = sim_res$power,
#    lower_CI = sim_res$lower_CI,
#    upper_CI = sim_res$upper_CI,
#    n_success = sim_res$n_reps
#    )

#  }, future.seed = TRUE)

# })
```

Adding results to prepared dataset
```{r}
# design_grid$power      <- sapply(results_list, `[[`, "power")
# design_grid$lower_CI   <- sapply(results_list, `[[`, "lower_CI")
# design_grid$upper_CI   <- sapply(results_list, `[[`, "upper_CI")
# design_grid$n_success  <- sapply(results_list, `[[`, "n_success")
```

Calculate increase [%] from RR
```{r}
# design_grid$perc_increase <- round((design_grid$RR - 1) * 100)
```

### Results of simulation

Loading saved results of simulation
```{r}
design_grid <- readRDS("Abundance_National.rds")
```

## Results visualisation

Substitute 1 for 0.999
```{r}
design_grid1 <- design_grid
if (any(design_grid1$power == 1, na.rm = TRUE)) {
  design_grid1$power[design_grid1$power == 1] <- 0.999
}
```

Final plot for abundance on national level
```{r fig1, fig.cap="Power analysis for abundance on national level"}
ggplot(design_grid1, aes(x = fish_per_pen, y = n_pens, z = power)) +
  geom_contour_filled(breaks = seq(0, 1, by = 0.1), alpha = 1) +
  geom_contour(breaks = 0.8,
               color = "black",
               linewidth = 1.1) +
  facet_grid(
    perc_increase ~ n_visits,
    labeller = labeller(
      n_visits = function(x)
        paste(x, "visits"),
      perc_increase = function(x)
        paste0(x, "% increase")
    )
  ) +
  scale_fill_viridis_d(name = "Power", option = "A", begin = 0.15, end = 0.9) +
  labs(
    x = "Number of fish sampled per pen",
    y = "Number of pens"
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), strip.text = element_text(face = "bold"))
```

Minimal sample size for each sampling scenario

```{r}
best_scenarios <- design_grid %>%
  mutate(No_Fish = fish_per_pen * n_visits * n_pens) %>%
  filter(power >= 0.8) %>%
  group_by(perc_increase, n_visits) %>%
  arrange(No_Fish, n_visits, n_pens, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  arrange(perc_increase, n_visits, n_pens, No_Fish) %>%
  as.data.frame()

Best<-best_scenarios[,c(9,3,4,2,10,5)]
Best[,6]<-paste0(round(best_scenarios$power,2),' [',round(best_scenarios$lower_CI,2),', ',
			round(best_scenarios$upper_CI,2),']')
```

```{r table1, results="asis", tab.cap="Minimal sample size for abundance at national level"}
knitr::kable(Best)
```

# Simulation based-power analysis on farm level

## Sampling scenarios

Farm level: 1 farm per 1 production year
```{r}
n_sites <- 1
n_years <- 1
```

Rates ratio (effect size)
```{r}
rate_ratios <- c(1.2, 1.3, 1.5, 2, 2.5)
```

Number of fish
```{r}
fish_vec <- c(5, 10, 15, 20, 25, 30)
```

Number of visits
```{r}
visit_vec <- c(4, 6, 8, 10, 12)
```

Number of pens
```{r}
pen_vec <- c(2, 5, 8, 10, 12, 15)
```

Replicates
```{r}
n_reps <- 1000
```

## Simulation run

In the following text the code to run simulation is provided, however, results are uploaded separately.

Create design grid for all scenarios
```{r}
design_grid <- expand.grid(
  RR = rate_ratios,
  fish_per_pen = fish_vec,
  n_visits = visit_vec,
  n_pens = pen_vec
)
```

Prepare columns for results
```{r}
design_grid$power <- NA_real_
design_grid$lower_CI <- NA_real_
design_grid$upper_CI <- NA_real_
design_grid$n_success <- NA_integer_
```

Set parallel backend
```{r}
# plan(multisession) 
```

Reproducibility
```{r}
# set.seed(123)
```

Progress bar
```{r}
# handlers(global = TRUE)
# handlers("txtprogressbar") 
```

Simulation
```{r}
#with_progress({
#
#  p <- progressr::progressor(steps = nrow(design_grid))

#  results_list <- future_lapply(seq_len(nrow(design_grid)), function(i) {
  
  # Convert RR to slope
#  slope_i <- log(design_grid$RR[i])
  
  # Run simulation
#  sim_res <- simulate_test_s(
#    site_sd = site_sd,
#    pen_sd = pen_sd,
#    year_sd = year_sd,
#    n_sites = n_sites,
#    n_pens = design_grid$n_pens[i],
#    n_years = n_years,
#    fish_per_pen = design_grid$fish_per_pen[i],
#    n_visits = design_grid$n_visits[i],
#    mean0 = mean0,
#    slope = slope_i,
#    theta = theta,
#    n_reps = n_reps
#  )

#    p()  
  
  # Return results
#  list(
#    power = sim_res$power,
#    lower_CI = sim_res$lower_CI,
#    upper_CI = sim_res$upper_CI,
#    n_success = sim_res$n_reps
#    )

#  }, future.seed = TRUE)

#})
```

Adding results to a prepared dataset
```{r}
# design_grid$power      <- sapply(results_list, `[[`, "power")
# design_grid$lower_CI   <- sapply(results_list, `[[`, "lower_CI")
# design_grid$upper_CI   <- sapply(results_list, `[[`, "upper_CI")
# design_grid$n_success  <- sapply(results_list, `[[`, "n_success")
```

Calculate percent increase from RR for clarity
```{r}
# design_grid$perc_increase <- (design_grid$RR - 1) * 100
```

### Results of simulation

Loading saved results of simulation
```{r}
design_grid <- readRDS("Abundance_Farm.rds")
```

## Results visualisation

Substitute 1 for 0.999 due to visualization
```{r}
design_grid1 <- design_grid
if (any(design_grid1$power == 1, na.rm = TRUE)) {
  design_grid1$power[design_grid1$power == 1] <- 0.999
}
```

Final plot for abundance on national level
```{r fig2, fig.cap="Power analysis for abundance at farm level"}
ggplot(design_grid1, aes(x = fish_per_pen, y = n_pens, z = power)) +
  geom_contour_filled(breaks = seq(0, 1, by = 0.1), alpha = 1) +
  geom_contour(breaks = 0.8,
               color = "black",
               linewidth = 1.1) +
  facet_grid(
    perc_increase ~ n_visits,
    labeller = labeller(
      n_visits = function(x)
        paste(x, "visits"),
      perc_increase = function(x)
        paste0(x, "% increase")
    )
  ) +
  scale_fill_viridis_d(name = "Power", option = "A", begin = 0.15, end = 0.9) +
  labs(
    x = "Number of fish sampled per pen",
    y = "Number of pens"
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), strip.text = element_text(face = "bold"))
```

Calculation of minimal sample size
```{r}
best_scenarios <- design_grid %>%
  mutate(No_Fish = fish_per_pen * n_visits * n_pens) %>%
  filter(power >= 0.8) %>%
  group_by(perc_increase, n_visits) %>%
  arrange(No_Fish, n_visits, n_pens, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  arrange(perc_increase, n_visits, n_pens, No_Fish) %>%
  as.data.frame()

Best<-best_scenarios[,c(9,3,4,2,10,5)]
Best[,6]<-paste0(round(best_scenarios$power,2),' [',round(best_scenarios$lower_CI,2),', ',
			round(best_scenarios$upper_CI,2),']')
```

```{r table2, results="asis", tab.cap="Minimal sample size for abundance at farm level"}
knitr::kable(Best)
```



